---
title: "Single-cell Immune Repertoire Trajectory Analysis Starting From scRepertoire"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{dandelionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
library(BiocStyle)
```

# Introduction

In a standard analysis workflow in R, users may choose to read in their VDJ data with [scRepertoire](https://github.com/ncborcherding/scRepertoire). In this vignette, we will demonstrate how to perform TCR trajectory analysis starting from 'raw' data i.e. just a standard single-cell gene expression data (stored in `SingleCellExperiment`) and VDJ data (in AIRR format).

## Installation

Install `scRepertoire` if you haven't already.
```{r, eval = FALSE}
if (!requireNamespace("scRepertoire", quietly = TRUE)) { # only for the tutorial
    BiocManager::install("scRepertoire")
}
# or
devtools::install_github("ncborcherding/scRepertoire")
```

### Load the required libraries
```{r}
library(dandelionR)
library(scRepertoire)
library(scater)
```

### Load the demo data
Due to size limitations of the package, we have provided an very trimmed down version of the demo data to ~2000 cells. The full dataset can be found here accordingly: GEX - https://developmental.cellatlas.io/fetal-immune (Lymphoid Cells) and VDJ - https://github.com/zktuong/dandelion-demo-files/tree/master/dandelion_manuscript/data/dandelion-remap

```{r}
data(demo_sce)
data(demo_airr)
```

### Use `scRepertoire` to load the VDJ data
For the trajectory analysis work here, we are focusing on the main productive TCR chains. Therefore we will flag `filterMulti = TRUE`, which will keep the selection of the 2 corresponding chains with the highest expression for a single barcode. For more details, refer to `scRepertoire`'s [documentation](https://www.borch.dev/uploads/screpertoire/reference/combinetcr).
```{r}
contig.list <- loadContigs(input = demo_airr, format = "AIRR")

# Format to `scRepertoire`'s requirements and some light filtering
combined.TCR <- combineTCR(contig.list,
    removeNA = TRUE,
    removeMulti = FALSE,
    filterMulti = TRUE,
    filterNonproductive = TRUE,
)
```

### Merging VDJ data with gene expression data
Next we will combine the gene expression data with the VDJ data to create a `SingleCellExperiment` object.
```{r}
sce <- combineExpression(combined.TCR, demo_sce)
```

### Initiate `dandelionR` workflow
Here, the data is ready to be used for the pseudobulk and trajectory analysis workflow in `dandelionR`.

Because this is a alpha-beta TCR data, we will set the `mode_option` to "abT". This will append `abT` to the relevant columns holding the VDJ gene information. If you are going to try other types of VDJ data e.g. BCR, you should set `mode_option` to "B" instead. And this argument should be consistently set with the `vdjPseudobulk` function later.

Since the TCR data is already filtered for productive chains in `combineTCR`, we will set `already.productive = TRUE` and can keep `allowed_chain_status` as `NULL`.

We will also subset the data to only include the main T-cell types: CD8+T, CD4+T, ABT(ENTRY), DP(P)_T, DP(Q)_T.
```{r}
sce1 <- setupVdjPseudobulk(sce, 
                          mode_option = "abT", 
                          already.productive = TRUE, 
                          subsetby = "anno_lvl_2_final_clean", 
                          groups = c("CD8+T", "CD4+T", "ABT(ENTRY)", "DP(P)_T", "DP(Q)_T"))
```

Visualise the UMAP of the filtered data.
```{r}
plotUMAP(sce, color_by = "anno_lvl_2_final_clean")
```


## Milo object and neighbourhood graph construction
We will use miloR to create the pseudobulks based on the gene expression data. The goal is to construct a neighbourhood graph with many neighbors with which we can sample the representative neighbours to form the objects.

```{r, warning = FALSE}
library(miloR)
milo_object <- Milo(sce)
milo_object <- buildGraph(milo_object, k = 30, d = 20, reduced.dim = "X_scvi")
milo_object <- makeNhoods(milo_object, reduced_dims = "X_scvi", d = 20, prop = 0.3)
```
### Construct UMAP on milo neighbor graph

We can visualise this milo object using UMAP.

```{r, warning = FALSE}
milo_object <- miloUmap(milo_object, n_neighbors = 30)
```
```{r}
plotUMAP(milo_object, color_by = "anno_lvl_2_final_clean", dimred = "UMAP_knngraph")
```

## Construct pseudobulked VDJ feature space

Next, we will construct the pseudobulked VDJ feature space using the neighbourhood graph constructed above. We will also run PCA on the pseudobulked VDJ feature space.
```{r}
pb.milo <- vdjPseudobulk(milo_object, mode_option = "abT", col_to_take = "anno_lvl_2_final_clean")

pb.milo <- runPCA(pb.milo, assay.type = "Feature_space", ncomponents = 20)
```

We can visualise the PCA of the pseudobulked VDJ feature space.
```{r}
plotPCA(pb.milo, color_by = "anno_lvl_2_final_clean")
```

## TCR trajectory inference using Absorbing Markov Chain

In the original `dandelion` python package, the trajectory inference is done using the `palantir` package. Here, we implement the absorbing markov chain approach in dandelionR to infer the trajectory, leveraging on `destiny` for diffusion map computation.

### Define root and branch tips
```{r}
# extract the PCA matrix
pca <- t(as.matrix(reducedDim(pb.milo, type = "PCA")))
# define the CD8 terminal cell as the top-most cell and CD4 terminal cell as the bottom-most cell
branch.tips <- c(which.max(pca[2, ]), which.min(pca[2, ]))
names(branch.tips) <- c("CD8+T", "CD4+T")
# define the start of our trajectory as the right-most cell
root <- which.min(pca[1, ])
```

### Construct diffusion map
```{r, warning = FALSE}
library(destiny)
# Run diffusion map on the PCA
dm <- DiffusionMap(t(pca), n_pcs = 10, n_eigs = 10)
```

### Compute diffussion pseudotime on diffusion map
```{r}
dif.pse <- DPT(dm, tips = c(root, branch.tips), w_width = 0.1)
```

```{r, message=FALSE}
# the root is automatically called DPT + index of the root cell
DPTroot <- paste0("DPT", root)
# store pseudotime in milo object
pb.milo$pseudotime <- dif.pse[[DPTroot]]
# set the colours for pseudotime
pal <- colorRampPalette(rev((RColorBrewer::brewer.pal(9, "RdYlBu"))))(255)
plotPCA(pb.milo, color_by = "pseudotime") + scale_colour_gradientn(colours = pal)
```

### Markov chain construction on the pseudobulk VDJ feature space

```{r}
pb.milo <- markovProbability(
    milo = pb.milo,
    diffusionmap = dm,
    terminal_state = branch.tips,
    root_cell = root,
    pseudotime_key = "pseudotime",
    knn = 30
)
```

### Visualising branch probabilities

With the Markov chain probabilities computed, we can visualise the branch probabilities towards CD4+ or CD8+ T-cell fate on the PCA plot.
```{r, message=FALSE}
plotPCA(pb.milo, color_by = "CD8+T") + scale_color_gradientn(colors = pal)
plotPCA(pb.milo, color_by = "CD4+T") + scale_color_gradientn(colors = pal)
```

## Transfer

The next step is to project the pseudotime and the branch probability information from the pseudobulks back to each cell in the dataset. If the cell do not belong to any of the pseudobulk, it will be removed. If a cell belongs to multiple pseudobulk samples, its value should be calculated as a weighted average of the corresponding values from each pseudobulk, where each weight is inverse of the size of the pseudobulk.

### Project pseudobulk data to each cell
```{r}
cdata <- projectPseudotimeToCell(milo_object, pb.milo, branch.tips)
```

### Visualise the trajectory data on a per cell basis
```{r, message=FALSE}
plotUMAP(cdata, color_by = "anno_lvl_2_final_clean", dimred = "UMAP_knngraph")
plotUMAP(cdata, color_by = "pseudotime", dimred = "UMAP_knngraph") + scale_color_gradientn(colors = pal)
plotUMAP(cdata, color_by = "CD4+T", dimred = "UMAP_knngraph") + scale_color_gradientn(colors = pal)
plotUMAP(cdata, color_by = "CD8+T", dimred = "UMAP_knngraph") + scale_color_gradientn(colors = pal)
```

And that's it! We have successfully inferred the trajectory of the T-cells in this dataset!

## Session info

```{r, warning = FALSE}
sessionInfo()
```


