---
title: "Single-cell Immune Repertoire Trajectory Analysis with dandelionR"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{dandelionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Foreword

Welcome to `dandelionR`!

`dandelionR` is an R package for performing single-cell immune repertoire trajectory analysis, based on the original python implementation in [dandelion](https://www.github.com/zktuong/dandelion). 

It provides all the necessary tools to interface with [scRepertoire](https://github.com/ncborcherding/scRepertoire) and a custom implementation of absorbing markov chain for pseudotime inference, inspired based on the [palantir](https://github.com/dpeerlab/Palantir) python package.

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Installation

You can install `dandelionR` from GitHub with:
```{r, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
devtools::install_github('tuonglab/dandelionR', dependencies = TRUE)
```

## Usage

```{r, message = FALSE, warning = FALSE}
library(dandelionR)
```

This is a work in progress, so please feel free to open an issue if you encounter any problems or have any suggestions for improvement.

## load data
load the demo data
```{r, eval = FALSE}
demo_pseudo_rds <- dandelionR::demo_pseudo_rds
```

## filter the data
subset the object to cells with paired chains, and prepare appropriately named and formatted columns for the pseudobulking functions to use as defaults
```{r, eval = FALSE}
demo_pseudo_rds <- setup_vdj_pseudobulk(demo_pseudo_rds, already.productive = FALSE)
```
plot the filtered data
```{r, eval = FALSE}
library(scater)
plotUMAP(demo_pseudo_rds,color_by = "anno_lvl_2_final_clean")
```

## format transferring

### milo object constructing and neighbour graph construction

## session info

```{r, message = FALSE, warning = FALSE}
sessionInfo()
```
