---
title: "Single-cell Immune Repertoire Trajectory Analysis with dandelionR"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{dandelionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Foreword

Welcome to `dandelionR`!

`dandelionR` is an R package for performing single-cell immune repertoire trajectory analysis, based on the original python implementation in [dandelion](https://www.github.com/zktuong/dandelion). 

It provides all the necessary tools to interface with [scRepertoire](https://github.com/ncborcherding/scRepertoire) and a custom implementation of absorbing markov chain for pseudotime inference, inspired based on the [palantir](https://github.com/dpeerlab/Palantir) python package.

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Installation

You can install `dandelionR` from GitHub with:
```{r, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
devtools::install_github('tuonglab/dandelionR', dependencies = TRUE)
```

## Usage

```{r, message = FALSE, warning = FALSE}
library(dandelionR)
```

This is a work in progress, so please feel free to open an issue if you encounter any problems or have any suggestions for improvement.

## load data
load the demo data
```{r, eval = FALSE}
data("demo_pseudo_rds", package = "dandelionR")
```

## filter the data
subset the object to cells with paired chains, and prepare appropriately named and formatted columns for the pseudobulking functions to use as defaults
```{r, eval = FALSE}
demo_pseudo_rds <- setup_vdj_pseudobulk(demo_pseudo_rds, already.productive = FALSE)
```
plot the filtered data
```{r, eval = FALSE}
library(scater)
plotUMAP(demo_pseudo_rds,color_by = "anno_lvl_2_final_clean")
```

## format transferring

### milo object constructing and neighbour graph construction

use miloR to create pseudobulks. Construct neighbor graph with many neighbors. Sample the representative neighbours form the objects.

```{r}
library(miloR)
traj_milo <- Milo(demo_pseudo_rds) 
milo_object <- buildGraph(traj_milo, k = 50, d=20, reduced.dim="X_scvi") 
milo_object <- makeNhoods(milo_object, reduced_dims="X_scvi", d=20)
```
## conduct umap on neighbor graph

```{r}
milo_object <- milo_umap(milo_object)
```
```{r}
scater::plotUMAP(milo_object,color_by = "anno_lvl_2_final_clean",dimred = "UMAP_knngraph")
```

### pseudobulk

construct vdj feature space

```{r}
cat("Column names in milo_object:\n")
library(SummarizedExperiment)
print("anno_lvl_2_final_clean" %in% names(colData(milo_object)))

```



```{r}
pb.milo <- dandelionR::vdj_pseudobulk(milo_object, col_to_take = "anno_lvl_2_final_clean")
```


```{r}

pb.milo <- scater::runPCA(pb.milo, assay.type = "X")
```

```{r}
plotPCA(pb.milo,color_by = "anno_lvl_2_final_clean")
```

### define root and branch tips
```{r}
library(SingleCellExperiment)
pca <- t(as.matrix(reducedDim(pb.milo, type = "PCA")))
branch.tips <- c(which.max(pca[2,]),which.min(pca[2,]))
names(branch.tips)<-c("CD8+T","CD4+T")
root <- which.min(pca[1,])
```

### make diffusion map
```{r}
library(destiny)
dm <- DiffusionMap(t(pca),n_pcs=50, n_eigs = 10) # have transition matrix in slot transitions
```

### compute diffussion pseudotime
```{r}
dif.pse <- DPT(dm, tips = c(root,branch.tips),w_width = 0.1)
```

```{r}
DPTroot <- paste0("DPT",root)
colors <- colorRampPalette(c("#03346E","#6EACDA", "white", "darkred"))(80)
plot(reducedDims(pb.milo)$PCA,  col = colors[cut(dif.pse[[DPTroot]],breaks=80)], pch=16, asp = 1)
library(fields)
image.plot(legend.only = TRUE, 
           zlim = range(dif.pse[[DPTroot]], na.rm = TRUE),  # 设置伪时间的范围
           col = colors,  # 与点颜色相同的渐变颜色条
           legend.lab = "Pseudotime")
```

## Markov chain construction, store the result in pb.milo

```{r}
diffusiontime <- dif.pse[[DPTroot]]
pb.milo <- markov_probability(pb.milo, dm, diffusiontime = diffusiontime, branch.tips, root)
```

```{r}
colors <- colorRampPalette(c("#03346E","#6EACDA", "white", "darkred"))(80)
plot(reducedDims(pb.milo)$PCA,  col = colors[cut(colData(pb.milo)[[names(branch.tips)[1]]],breaks=80)], pch=16, asp = 1)

image.plot(legend.only = TRUE, 
           zlim = range(colData(pb.milo)[[names(branch.tips)[1]]], na.rm = TRUE),  
           col = colors,  
           legend.lab = "Pseudotime")  
plot(reducedDims(pb.milo)$PCA,  col = colors[cut(colData(pb.milo)[[names(branch.tips)[2]]],breaks=80)], pch=16, asp = 1)

image.plot(legend.only = TRUE, 
           zlim = range(colData(pb.milo)[[names(branch.tips)[2]]], na.rm = TRUE),  
           col = colors,  
           legend.lab = "Pseudotime") 
```

# Transfer

### project pseudobulk data to each cell
```{r}
cdata<- project_pseudotime_to_cell(milo_object, pb.milo, branch.tips)
```
```{r}
plotUMAP(cdata,color_by = "anno_lvl_2_final_clean",dimred = "UMAP_knngraph")
plotUMAP(cdata,color_by = "pseudotime",dimred = "UMAP_knngraph")
plotUMAP(cdata,color_by = "CD4+T",dimred = "UMAP_knngraph")
plotUMAP(cdata,color_by = "CD8+T",dimred = "UMAP_knngraph")
```


## session info

```{r, message = FALSE, warning = FALSE}
sessionInfo()
```
