---
title: "dandelionR"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{dandelionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# dandelionR

Welcome to `dandelionR`! `dandelionR` is an R package for performing single-cell immune repertoire trajectory analysis, based on the original python implementation in [`dandelion`](https://www.github.com/zktuong/dandelion). It provides all the necessary tools to interface with [`scRepertoire`](https://github.com/ncborcherding/scRepertoire) and a custom implementation of absorbing markov chain for pseudotime inference, inspired based on the [palantir](https://github.com/dpeerlab/Palantir) python package.

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Installation

You can install `dandelionR` from GitHub with:
```{r, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
devtools::install_github('tuonglab/dandelionR', dependencies = TRUE)
```

## Usage

```{r, message = FALSE, warning = FALSE}
library(dandelionR)
```

This is a work in progress, so please feel free to open an issue if you encounter any problems or have any suggestions for improvement.
